% Cover Letter for Nature Machine Intelligence submission
% All authors anonymized

XXX
XXX, XXX, XXX

Dear Editor,

We submit our manuscript entitled "Spike-Trace as a World State: Pure-SNN JEPA World Models" for consideration at Nature Machine Intelligence.

The paper introduces ST-JEWM, a pure-SNN (no Transformer, no attention) reconstruction-free world model that uses a content-aware gated spike trace as its world-state interface. On the standard DMC Reacher physical-task bench (real mujoco 3.10 simulation, $n{=}50$ episodes), our best model v4.5 trained on mujoco rollouts achieves a $14\%$ success rate, $+8$ percentage points above a LeWM-style Transformer baseline trained on the same data. The architecture is competitive with the LeWM baseline (which uses a Transformer) while being a pure-SNN model with $5.03$M trainable parameters (vs LeWM's $18.77$M, $0.27{\times}$ the parameter count).

We provide:
\begin{itemize}
    \item Three theoretical propositions (Trace Boundedness, Gate Stability, Loss Monotonicity) with complete proofs and Monte Carlo validation ($N{=}10{,}000$, all PASS).
    \item A real mujoco 3.10 physical benchmark result that outperforms a LeWM-style Transformer baseline.
    \item A clear training-data insight: the static dataset's $qpos \equiv 0$ prevents learning $(qpos, action) \to qpos'$ dynamics; training on mujoco rollouts (where $qpos$ varies) unlocks the physical-task win.
    \item A negative result on two-stage training that demonstrates the value of joint training.
\end{itemize}

The spike-trace world-state interface decouples the membrane potential (a leaky continuous variable) from the world-state read-out (a content-aware gated trace of discrete spike events). This is a natural fit for event-driven neuromorphic hardware (Loihi, Akida) where sparsity maps directly to energy efficiency. The three propositions in this paper provide a theoretical foundation for these downstream applications.

We have no conflicts of interest to declare. We suggest the following potential reviewers:
\begin{itemize}
    \item XXX — SNN expert
    \item XXX — World models expert
    \item XXX — Energy-efficient ML expert
\end{itemize}

We look forward to your response.

Sincerely,
XXX
